```
title: "CLOPE Example"
```

Load the CLOPE R source code.

```{r}
source("clope.R")
```

# Simple example

Create example data. Can start as data frame.

```{r}
basket_df <- data.frame(
  ID = c(1,1,1,2,2,3,3,3,4,4,4,5,5),
  Item = c("milk","bread","eggs","bread","butter","milk","bread","butter",
           "milk","diapers","beer","bread","milk"),
  Number =   c(1,2,1,2,1,1,1,1,2,1,2,3,1)
)
basket_df
```

Or a list of named numeric vectors.

```{r}
#| eval: false
basket_list <- list(
  c("milk"=1, "bread"=2, "eggs"=1),
  c("bread"=2, "butter"=1),
  c("milk"=1, "bread"=1, "butter"=1),
  c("milk"=2, "diapers"=1, "beer"=2),
  c("bread"=3, "milk"=1)
)
```

Convert data to a list of named numeric vectors if it is a data frame.

```{r}
basket_items <- split(basket_df[c("Item","Number")], basket_df$ID)
basket_list <- lapply(basket_items, \(df) {
  x <- df$Number
  names(x) <- df$Item
  x
})
```

Run CLOPE.

```{r}
clope(basket_list)
```

# Bigger example

Load data.

```{r}
library(arules)
data("Groceries")
```

Convert arules transaction data to a list of named numeric vectors.

```{r}
tx_list <- as(Groceries, "list")
any(unlist(lapply(tx_list, duplicated))) # Check if there are any duplicates
tx <- lapply(tx_list, \(items) {
  count <- rep(1, length(items))
  names(count) <- items
  count
})
```

Run CLOPE. This will take a while.

```{r}
tx_clust <- clope(tx)
```

Now do something with the clustered data.
